Trajectories

### 2023-01-17 docs/index.qmd
Update lab notes
Add figures to manuscript, start discussion, add model and predictor appendix

---
title: "Trajectories"
authors:
    - Gonzalo Garcia-Castro^[Universitat Pompeu Fabra, gonzalo.garciadecastro@upf.edu]
    - Daniela Avila-Varela^[Universitat Pompeu Fabra]
    - Ignacio Castillejo^[Universidad Autónoma de Madrid]
    - Núria Sebastian-Galles^[Universitat Pompeu Fabra]
output:
    html_document:
        code_folding: hide
        number_sections: true
        toc: true
    github_document:
        pandoc_args: --webtex
        number_sections: true
        toc: true
bibliography: "../references.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    message = TRUE,
    echo = TRUE,
    warning = TRUE,
    include = TRUE,
    out.width = "80%",
    dpi = 250
)
options(
    knitr.kable.NA = "-",
    knitr.duplicate.label = "allow",
    ggplot2.discrete.fill = ggsci::pal_d3()(4),
    ggplot2.discrete.colour = ggsci::pal_d3()(4),
    ggplot2.continuous.fill = ggplot2::scale_color_gradient,
    ggplot2.continuous.colour = ggplot2::scale_color_gradient
)

>>>>

---
title: "Trajectories"
subtitle: "Lab notes"
date: "Updated: `r format(Sys.Date(), '%Y/%m/%d')`"
authors:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    corresponding: true
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias fargas 25-27
        city: Barcelona
        state: Spain
        postal-code: 08005
  - name: Daniela S. Ávila-Varela
    orcid: 0000-0002-3518-8117
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations:
      - name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
abstract: "Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those in the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months."
format:
    html:
        code-copy: hover
        reference-location: margin
        code-overflow: scroll
    pdf: default
link-external-newwindow: true
code-fold: true
code-line-numbers: true
fig-dpi: 1000
warning: false
echo: false
toc: true
colorlinks: true
number-sections: true
bibliography: "references.bib"
csl: "apa7.csl"
---

```{r}
#| label: setup
#| include: false

# import packages
library(conflicted)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(forcats)
library(stringr)
library(tibble)
library(tidybayes)
library(janitor)
library(brms)
library(scales)
library(marginaleffects)
library(lubridate)
library(bvqdev)

# load targets as R objects
tar_load(participants)
tar_load(df)
tar_load(items)
tar_load(bvq_data)
tar_load(model_fit_0)
tar_load(model_fit_1)
tar_load(model_fit_2)
tar_load(model_fit_3)
tar_load(model_fit_4)
tar_load(model_fit_4_prior)
tar_load(model_fit_prior)
tar_load(model_loos)
tar_load(rope_interval)
tar_load(posterior_description)
tar_load(posterior_description_re)

# resolve namespace conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

# set ggplot theme and colour palette
theme_set(theme_custom()) # set custom ggplot theme
clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = ggplot2::scale_color_gradient,
        ggplot2.continuous.colour = ggplot2::scale_color_gradient)
```

# Methods

::: {.callout-tip}
Data and materials are available at [OSF](https://osf.io/hy984/), and code is available at [GitHub](https://github.com/gongcastro/trajectories).
:::

## Questionnaires

```{r}
#| label: questionnaire
#| echo: false
#| message: false
#| warning: false
data("pool", package = "bvqdev")
n_categories <- distinct(pool, semantic_category) %>% nrow()
n_items <- length(unique(pool$item))
n_item_language <- count(pool, language) %>% 
    pull(n) %>%
    set_names(c("catalan", "spanish"))
```

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted of two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]` items. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* \[dog\] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* \[*to finish*\] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* \[*monkey*\] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (800 translation equivalents; 800 items in Catalan, 800 items in Spanish), while others filled a shorter version (\~400 translation equivalents, \~400 items in Catalan, \~400 items in Spanish). These last families were randomly assigned to one of four different subsets of the complete list of items. These lists were designed so that each contained a representative subsample of the items from the complete list. Semantic/functional categories with less than 16 items were not divided in the short version of the questionnaire to preserve a minimum of four items per list for each category: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions. We excluded from the analysis multi-word items (e.g., *barrita de cereales* \[cereal bar\]) and items that included more than one word-form (e.g., *mono / mico*). Table 3 shows the classification of items in cognates and non-cognates and their lexical frequency scores across the four lists of the inventories.

## Participants

```{r}
#| label: fig-participants-time
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 10 
#| fig-cap: "Date at which responses were collected"
participants %>% 
    group_by(time_stamp) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    mutate(n = cumsum(n)) %>% 
    ggplot(aes(time_stamp, n)) +
    geom_vline(xintercept = ymd("2020-03-15"), size = 0.5) +
    annotate(geom = "text", y = 350, x = ymd("2020-03-15"), label = "COVID-19 lockdown in Spain",
             angle = 90, vjust = 1.5, hjust = 1) +
    geom_line(size = 1, colour = clrs[1]) +
    labs(x = "Date", y = "Number of responses", colour = "Dominant language") +
    scale_y_continuous(breaks = seq(0, 500, 100), limits = c(0, 500)) +
    theme(legend.position = "top",
          legend.title = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 20, hjust = 1),
          axis.title.x = element_blank()) +

    participants %>% 
    mutate(age = cut(floor(age), breaks = seq(7, 35, 5))) %>% 
    group_by(time_stamp, age) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    group_by(age) %>% 
    mutate(n = cumsum(n)) %>% 
    ggplot(aes(time_stamp, n, colour = age)) +
    geom_vline(xintercept = ymd("2020-03-15"), size = 0.5) +
    geom_line(size = 1) +
    labs(x = "Date", y = "Number of responses", colour = "Age (months(") +
    scale_y_continuous(breaks = seq(0, 500, 50)) +
    scale_colour_manual(values = clrs) +

    plot_layout(ncol = 1) &

    scale_x_date(date_breaks = "3 month", date_labels = "%B %Y") &
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1),
        axis.title.x = element_blank(),
    )
```

### Age

Number of months elapsed between participants' birth date and questionnaire completion.

```{r}
#| label: fig-participants-age
#| fig-width: 7
#| fig-cap: "Distribution of participants' ages"
participants %>% 
    mutate(age = floor(age)) %>% 
    count(age) %>% 
    ggplot() +
    aes(x = age, y = n) +
    geom_col(fill = clrs[1]) +
    geom_text(aes(label = n), size = 3.5, vjust = -1) +
    labs(
        x = "Age (months)",
        y = "# participants"
    ) +
    scale_x_continuous(breaks = seq(0, 40, 2)) +
    scale_y_continuous(limits = c(0, 55), breaks = seq(0, 55, 10)) +
    theme(
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()
    )
```

### Language exposure (Degree of Exposure)

Participant degree of exposure to the words' language ($DoE$): percentage of exposure to the language the word belonged to. For example, for a participant with 90% exposure to Catalan, and 10% to Spanish, the DoE of the Catalan word *taula* would be 90%, and the DoE for the Spanish word *mesa* would be 10%.

```{r}
#| label: fig-participants-lp
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Participant degree of exposure (DoE) to Catalan and Spanish"
participants %>% 
    pivot_longer(
        c(doe_catalan, doe_spanish),
        names_to = "language", values_to = "doe", 
        names_transform = ~str_to_sentence(str_remove(., "doe_"))
    ) %>% 
    mutate(
        doe = cut(doe, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
        id = paste0(id, " (", time, ")")
    ) %>% 
    count(language, doe) %>% 
    ggplot(aes(doe, y = n, label = n, fill = language)) +
    geom_col(colour = "white", size = 1) +
    geom_text(position = position_nudge(y = 5)) +
    facet_wrap(~language, ncol = 1) + 
    labs(
        x = "Degree of exposure to Catalan",
        y = "Number of participants",
        colour = "Language",
        fill = "Language"
    ) +
    scale_fill_manual(values = clrs[c(1, 4)]) +
    theme(
        legend.position = "top",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()
    )
```

### SES/parental education

<!-- # ```{r} -->
<!-- #| label: tbl-participants-edu -->
<!-- #|  tbl-cap: "Participant exposure to the language (DoE) and socio-economic status (SES)." -->
<!-- # participants %>% -->
<!-- #     left_join(bvq_data$logs) %>% -->
<!-- #     mutate( -->
<!-- #         edu_parent = fct_explicit_na(as.factor(edu_parent)), -->
<!-- #         doe_catalan = cut(doe_catalan, breaks = seq(0, 1, 0.2), include.lowest = TRUE) -->
<!-- #     ) %>% -->
<!-- #     count(doe_catalan, edu_parent) %>% -->
<!-- #     right_join( -->
<!-- #         expand_grid( -->
<!-- #             doe_catalan = levels(.$doe_catalan), -->
<!-- #             edu_parent = unique(bvq_data$logs$edu_parent) -->
<!-- #         ) -->
<!-- #     ) %>% -->
<!-- #     rename(total = n) %>% -->
<!-- #     pivot_wider(names_from = edu_parent, -->
<!-- #                 values_from = total,  -->
<!-- #                 values_fill = 0) %>% -->
<!-- #     mutate(total = rowSums(cbind(.[,3:8]), na.rm = TRUE)) %>% -->
<!-- #     rename(Total = total) %>% -->
<!-- #     select(-"NA") %>%  -->
<!-- #     gt()  -->
<!-- # summary_rows( -->
<!-- #     columns = Complementary:Primary, -->
<!-- #     fns = list(Total = ~sum(., na.rm = TRUE)), -->
<!-- #     formatter = fmt_integer -->
<!-- # ) %>% -->
<!-- # cols_label(doe_catalan = "DoE Catalan") %>% -->
<!-- # sub_missing(columns = 2:8, missing_text = "--") %>% -->
<!-- # tab_spanner("Educational attainment", Secondary:Primary) %>% -->
<!-- # tab_style( -->
<!-- #     style = list(cell_borders(weight = px(2), sides = "left", color = "grey50")), -->
<!-- #     locations = cells_body(columns = 9, rows = 1:5) -->
<!-- # ) %>% -->
<!-- # tab_style( -->
<!-- #     cell_text(style = "italic"), -->
<!-- #     cells_column_labels(columns = 2:7) -->
<!-- # ) %>% -->
<!-- # tab_style( -->
<!-- #     cell_text(weight = "bold"), -->
<!-- #     cells_column_labels(columns = c(1, 8)) -->
<!-- # ) %>% -->
<!-- # tab_style( -->
<!-- #     cell_text(weight = "bold"), -->
<!-- #     cells_column_spanners(spanners = "Educational attainment") -->
<!-- # ) -->
<!-- ``` -->

## Items

```{r}
#| label: tbl-items-summary
#| tbl-cap: "Lexical frequencies. Mean, standard error, and 95% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."
# items <- left_join(items, select(bvq_data$pool, te, item, language, class, semantic_category))
# 
# items %>%
#     unnest(list) %>%
#     group_by(language, list) %>%
#     summarise_at(
#         vars(freq_zipf, n_phon, lv),
#         list(
#             n = ~sum(!is.na(.)),
#             mean = ~mean(., na.rm = TRUE),
#             sd = ~sd(., na.rm = TRUE),
#             min = ~min(., na.rm = TRUE),
#             max = ~max(., na.rm = TRUE)
#         )
#     ) %>%
#     select(-c(n_phon_n, lv_n)) %>%
#     rename(n = freq_zipf_n) %>%
#     gt(groupname_col = "language", rowname_col = "list") %>%
#     fmt_number(matches("freq_|n_phon_mean|n_phon_sd"), decimals = 1) %>%
#     fmt_percent(matches("lv_"), decimals = 1) %>%
#     tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>%
#     tab_spanner("Length (Phonemes)", columns = starts_with("n_phon_")) %>%
#     tab_spanner("Phon. similarity (Levenshtein)", columns = starts_with("lv_")) %>%
#     cols_merge_uncert(col_val = freq_zipf_mean, col_uncert = freq_zipf_sd) %>%
#     cols_merge_uncert(col_val = n_phon_mean, col_uncert = n_phon_sd) %>%
#     cols_merge_uncert(col_val = lv_mean, col_uncert = lv_sd) %>%
#     cols_merge_range(col_begin = freq_zipf_min, col_end = freq_zipf_max) %>%
#     cols_merge_range(col_begin = n_phon_min, col_end = n_phon_max) %>%
#     cols_merge_range(col_begin = lv_min, col_end = lv_max) %>%
#     cols_label(
#         list = "",
#         n = md("*N*"),
#         freq_zipf_mean = md("Mean ± SD"),
#         freq_zipf_min = md("Range"),
#         n_phon_mean = md("Mean ± SD"),
#         n_phon_min = md("Range"),
#         lv_mean = md("Mean ± SD"),
#         lv_min = md("Range")
#     ) %>%
#     tab_style(
#         cell_text(style = "italic"),
#         cells_column_labels(everything())
#     ) %>%
#     tab_style(
#         cell_text(weight = "bold"),
#         cells_column_spanners(everything())
#     ) %>%
#     summary_rows(
#         fns = list("Grand mean" = ~mean(.)),
#         columns = ends_with("_mean")
#     ) %>%
#     summary_rows(
#         fns = list("N" = ~sum(.)),
#         columns = n,
#         formatter = fmt_integer
#     )
```

### Lexical frequency

Word lexical frequency ($Frequency$): lexical frequency of the word in its corresponding language, expressed as Zipf scores [@van2014subtlex; @zipf1949human]. This variable ranges from 0 to 7, and follows and approximates a normal distribution, with most values in a corpus ranging from 3 to 5 points. Lexical frequencies were extracted from the English corpora of the [CHILDES](https://childes.talkbank.org/) database using the [{childesr}](https://langcog.github.io/childes-db-website/) R package. Available words in the corpora were then mapped to their Catalan and Spanish translations. Therefore, the Catalan and Spanish word-forms of the same translation equivalent had the same value for lexical frequency. Responses to words with missing lexical frequencies were excluded from analyses of the total number of items).

```{r}
#| label: fig-items-freq
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Distribution of lexical frequency across items. Lexical frequencies were extracted from the English CHILDES corpora, and mapped to their Catalan and Spanish translations. Therefore Catalan and Spanish word-forms that belong to the same translation equivalent had the same lexical frequency. The black dot and interval indicate the mean and standard deviation."
mean_sd <- function(x, ...) {
    m <- mean(x, ...)
    s <- sd(x, ...)
    data.frame(mean = m, ymin = m-s, ymax = m+s)
}

items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    ggplot(aes(x = freq_zipf)) +
    geom_histogram(fill = clrs[4], colour = "white", bins = 30)  +
    labs(x = "Lexical frequency (Zipf score)", y = "Number of words") +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        legend.title = element_blank()
    )
```

### \# Phonemes

Word length ($Phonemes$), computed as the number of phonemes in the phonological transcription of the word-form in [SAMPA](https://en.wikipedia.org/wiki/SAMPA) format.

```{r}
#| label: fig-items-nphon
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of number of phonemes across items"
items %>%
    mutate(language = ifelse(str_detect(item, "cat_"), "Catalan", "Spanish")) %>%
    count(language, n_phon) %>%
    ggplot() +
    aes(
        x = n_phon,
        y = n,
        fill = language,
        colour = language,
        shape = language
    ) +
    facet_wrap(~language, ncol = 1) +
    geom_col(colour = "white") +
    geom_text(aes(label = n), position = position_nudge(y = 5), colour = "black") +
    labs(
        x = "# Phonemes",
        y = "Number of words",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_x_continuous(breaks = 1:15) +
    scale_fill_manual(values = clrs[c(1, 4)]) +
    theme(
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.x = element_line(),
        legend.title = element_blank()
    )
```

### Phonological similarity (Levenshtein)

Phonological similarity between translation equivalents was computed as the normalised Levenshtein similarity between the word-form and its translation in the other language [@levenshtein1966binary]. This score is calculated by first calculating the Levenshtein distance between the two transcriptions (number of insertions, deletions or replacements needed for the shortest transcription to become identical to the longer transcription), then dividing the resulting value by the length of the longest transcription, and finally subtracting this value from 1. This results in a proportion that indicates how much the two phonological transcriptions of the translation equivalent are similar to each other, ranging from 0% (no similarity at all) to 100% (both transcriptions are identical) [see @floccia2018vocabulary; @fourtassi2020growth; @laing2022phonological for similar approaches].

```{r}
#| label: fig-items-lv
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-cap: "Distribution of Levenshtein similarities across items"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    mutate(lv = cut(lv, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>% 
    count(lv) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = lv, n)) +
    geom_col(fill = clrs[4], colour = "white") +
    geom_text(aes(label = paste0(n, " (", percent(prop, accuracy = 0.1), ")")),
              position = position_nudge(y = 3), size = 3.5) +
    labs(x = "Levenshtein distance", y = "Number of words") +
    theme(
        panel.grid.major.x = element_blank(),
        legend.position = "top"
    )
```

### Frequency by \# Phonemes

```{r}
#| label: fig-items-frequency-phonemes
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Association betwen number of lexical frequency and number of phonemes"
items %>% 
    drop_na(freq_zipf) %>% 
    distinct(te, .keep_all = TRUE) %>% 
    count(n_phon, freq_zipf) %>% 
    ggplot(aes(n_phon, freq_zipf)) +
    geom_point(size = 2, alpha = 0.5, colour = clrs[4], shape = 1, stroke = 1,
               position = position_jitter(width = 0.1), seed = 888) +
    geom_smooth(method = "lm", formula = "y ~ x" ,
                colour = clrs[1], fill = clrs[1]) +
    labs(
        x = "# Phonemes", 
        y = "Lexical frequency (Zipf score)", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

### Frequency by Levenshtein

```{r}
#| label: fig-items-frequency-lv
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen lexical frequency and phonological similarity"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    drop_na(freq_zipf) %>% 
    count(freq_zipf, lv) %>% 
    ggplot(aes(freq_zipf, lv, size = n)) +
    geom_point(size = 2, shape = 1, stroke = 1, colour = clrs[4]) +
    geom_smooth(method = "lm", formula = "y ~ x", colour = clrs[1], fill = clrs[1]) +
    labs(
        y = "Levenshtein similarity", 
        x = "Lexical frequency (Zipf score)", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid.major.y = element_blank()
    )
```

### \# Phonemes by Levenshtein

```{r}
#| label: fig-items-nphon-lv
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association between number of phonemes and phonological similarity"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, lv) %>% 
    ggplot(aes(n_phon, lv, size = n)) +
    geom_point(shape = 1, stroke = 1, colour = clrs[4]) +
    geom_smooth(method = "lm", formula = "y ~ x", size = 1,
                colour = clrs[1], fill = clrs[1]) +
    labs(
        x = "# Phonemes", 
        y = "Levenshtein similarity", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

### Word exposure index by lexical frequency

We created a predictor ($Exposure$) to account for the exposure rate of each child to each of the words their parents responded to, weighted by the child's exposure to the language the word belongs to. the exposure of the $i$-th child to the $j$-th word measure is the product of the lexical frequency of the word (Zipf score) and the child’s degree of exposure to the corresponding language (a proportion) (see @eq-exposure).

$$
\textbf{Exposure}_{ij} = \textbf{Frequency}_i \times \textbf{DoE}_{j}
$$ {#eq-exposure}

For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure to the word *cavall* (horse, in Catalan, with a lexical frequency of `r round(items$freq_zipf[items$item=="cat_cavall"], 2)`) would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.20, 2)` (see @eq-exposure-example):

$$
\begin{aligned}
\textbf{Exposure}_{\textbf{cavall, Catalan}} &= \textbf{Frequency}_{cavall} \times \textbf{DoE}_{Catalan} = 6.14 \times 0.8 = 4.91 \\
\textbf{Exposure}_{\textbf{caballo, Spanish}} &= \textbf{Frequency}_{caballo} \times \textbf{DoE}_{Spanish} = 6.14 \times 0.2 = 1.23
\end{aligned}
$$ {#eq-exposure-example}

```{r}
#| label: fig-exposure-distribution
#| fig-cap: "Relationship between lexical frequency and language exposure-weighted lexical frequency"
#| warnings: false
#| fig-height: 7
#| fig-width: 13
df %>%
    mutate(
        language = ifelse(grepl("cat_", item), "Catalan", "Spanish"),
        exposure = cut(
            exposure,
            breaks = seq(0, 9, 1),
            include.lowest = TRUE
        )
    ) %>%
    count(exposure, language) %>%
    ggplot(aes(exposure, n, fill = language)) +
    facet_wrap( ~ language) +
    geom_col(colour = "white") +
    labs(x = "Exposure-weighted frequency\n(Frequency \u00d7 DoE)",
         y = "Number of observations",
         fill = "Language") +

    df %>%
    mutate(
        language = ifelse(grepl("cat_", item), "Catalan", "Spanish"),
        exposure = cut(
            exposure,
            breaks = seq(0, 9, 1),
            include.lowest = TRUE
        ),
        freq = cut(freq, breaks = seq(0, 9, 0.25), include.lowest = TRUE)
    ) %>%
    count(exposure, freq, language) %>%
    ggplot(aes(
        x = freq,
        y = exposure,
        colour = language,
        size = n
    )) +
    facet_wrap( ~ language) +
    geom_point() +
    labs(
        x = "Frequency (Zipf score)",
        y = "Exposure-weighted frequency\n(Frequency \u00d7 DoE)",
        colour = "Language",
        fill = "Language",
        size = "Number of observations"
    ) +
    theme(panel.grid = element_blank()) +
    plot_layout(ncol = 1) &
    scale_colour_manual(values = clrs[c(1, 4)]) &
    scale_fill_manual(values = clrs[c(1, 4)]) &
    guides(fill = "none", colour = "none") &
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.y = element_blank())
```

## Data analysis

We used multilevel ordinal regression to model the cumulative probability of a *No* response, a *Understands* response, or a *Understands and Says* response [@burkner2019ordinal] using the *logit* link function. The likelihood of the cumulative probability distribution of the responses is defined by @eq-likelihood (see @fig-likelihood for a graphical representation).

$$
\begin{aligned}
\textbf{Likelihood:} \\
y_{ij} &\sim \text{Cumulative}(p_{k})
\end{aligned}
$$ {#eq-likelihood}

where:

- $y$ is an observed response ($y \in \{\text{No, Understands, Understands and Says}\}$)
- $i$ is the participant index
- $j$ is the translation equivalent (TE) index
- $p_{k}$ is a probability ($p \in (0, 1)$) that indicates the threshold $k$ ($k \in (1, 2)$)) between two response categories in the latent distribution
$p_{k}$ is then estimated using a logit regression model as indicated in @eq-linear.

```{r}
#| label: fig-likelihood
#| fig-cap: "Graphical representation of the assumed latent distribution that generates observed responses in the questionnaire. Cumulative ordinal regression assumes that observed categorical responses (*No*, *Understands* and *Understands and Says* are generated from a common latent (i.e., unobserved) distribution. For threotical reasons, we refer to the dimension captured by such distribution as *Familiarity*. *Familiarity* describes a given participant's familiarity with a given word. When familiarity is low (below some threshold 1), the reulting response is *No*. When familiarity is higher than such threshold 1, but lower than some threshold 2, the resulting result is *Understands* (i.e., comprehension). When familiarity exceeds threshold 2, the resulting response is *Understands and Says* (i.e., comprehension and production). These thresholds are estimated in the cumulative regression model as intercepts 1, and 2. The rest of the coefficients in the regression model shift the response thresholds, making comprehension or production more likely or less likely. See @burkner2019ordinal for a detailed description of cumulative ordinal regression and its implementation."
#| fig-width: 7
#| fig-height: 3.5
dist_df = tribble(~mean, ~sd, 0,   1)

responses <- data.frame(
    ymin = c(-Inf, -1.5, 1.5),
    ymax = c(-1.5, 1.5, Inf),
    xmin = -Inf,
    xmax = Inf,
    .category = str_wrap(c("No", "Understands", "Understands and Says"), width = 12)
)

responses %>%
    ggplot(aes(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, fill = .category)) +
    geom_rect(alpha = 0.5) +
    stat_dist_slab(
        data = dist_df, 
        aes(dist = "norm", arg1 = mean, arg2 = sd),
        position = "dodge", inherit.aes = FALSE,
        fill = NA, colour = "black"
    ) +
    geom_hline(yintercept = c(-1.5, 1.5), linetype = "dotted", size = 1) +
    geom_text(aes(label = .category, y = c(-3, 0, 3), x = 1),
              vjust = 1, hjust = 0.5) +
    annotate(
        geom = "text",
        label = c("Threshold~1~(p[1])", "Threshold~2~(p[2])"),
        y = c(-1.5, 2), parse = TRUE,
        x = 0.5, angle = 90, vjust = -0.5
    ) +
    labs(y = "Hypothetical response-generating latent distribution",
         x = "Response probability",
         fill = "Response type",
         colour = "Response type") +
    coord_flip() +
    scale_fill_manual(values = clrs[c(1, 4, 5)]) +
    scale_x_continuous(labels = percent) +
    scale_y_continuous(limits = c(-4, 4)) +
    theme(panel.grid = element_blank(),
          legend.position = "none")
```

To test our hypotheses, we included several predictors in the regression model as fixed effects: the main effects of participant age ($Age$), number of phonemes ($Phonemes$), word exposure index ($Exposure$), and of phonological similarity ($Levenshtein$), the two-way interactions $Age \times Exposure$, $Age \times Levenshtein$, and $Exposure \times Leveshtein$, and the three-way interaction $Age \times Exposure \times Leveshtein$. We also included crossed random effects for participants and translation equivalents to account for the repeated measures in our dataset--each participant provided responses to multiple translation equivalents, and each translation equivalent was responded to by multiple participants [@gelman2020regression]. For both grouping variables, we included random intercepts, random slopes, and correlation parameters for all predictors were repeated measures were observed in our dataset [@barr2013random, see @eq-linear].

$$
\begin{aligned}
\textbf{Linear model:} \\
logit(p_{k}) = \text{ln} \frac{p_{k}}{1-p_{k}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + \\
& (\beta_{1} + u_{1_{i}} + w_{1_{j}}) · \text{Age}_{i} + & \\
& (\beta_{2} + u_{2_{i}} + w_{2_{j}}) · \text{Phonemes}_{ij} + & \\
& (\beta_{3} + u_{3_{i}} + w_{3_{j}}) · \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) · \text{Levenshtein}_{ij} + & \\
& (\beta_{5} + u_{5_{i}} + w_{3_{j}}) · (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) · (\text{Age}_{i} \times \text{Levenshtein}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) · (\text{Exposure}_{ij} \times \text{Levenshtein}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) · (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Levenshtein}_{ij}) & \\
\end{aligned}
$$ {#eq-linear}

where:

- $i$ and $j$ index the participant and translation equivalent (TE)
- $\beta_{0_k}$ is the fixed coefficient of the regression model for the intercept of threshold $k$
- $u_{0_{i}}$ and $w_{0_{j}}$ are the by-participant and by-TE adjustments for $\beta_{0_{k}}$ (i.e., random intercepts), respectively
- $\beta_{1-8}$ are the fixed coefficients of the regression model for the predictors of interest
- $u_{1-8_{i}}$ and $w_{1-3_{j}}$ are the by-participant and by-TE adjustments for$\beta_{1-8}$ (i.e., random slopes), respectively

We used the Bayesian framework to estimate the parameters in our model. This involves using the Bayes theorem to compute a distribution (*posterior distribution*) that describes what values of each parameter in the model are more likely given the data (*likelihood*, see @eq-likelihood), and previous knowledge about such distribution (*prior distribution*, see @eq-prior) [@mcelreath2020statistical]. This posterior distribution not only informs about the most likely values of our regression coefficients of interest, but also about the uncertainty around such estimations. We used a weakly informative prior for our parameters, with the exception of the main effect of $Age$, for which we specified a strongly informative prior based on previous literature about how age affects the acquisition of words [see @eq-prior].
